---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `brfss2013`. Delete this note when before you submit 
your work. 

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

Genetics,  
* * *

## Part 2: Research questions

**Research quesion 1:**

Is the level of education of working people lead to the higher income?

**Research quesion 2:**

Is the level of education of working native americans/alaskian people's lead to the higher income?

**Research quesion 3:**

Is childfree educated womens become to drink more alcohol with ages?

**Research quesion 4:**

Is self employed mans who sleeps less than it is necessary, drink more alcohol?

* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1:**
Is the level of education of working people lead to the higher income?

At first lets take a look at the BRFSS's Codebook and try to achieve what variables can help us with this investigation.  

```{r}

```



**Research quesion 2:**

```{r}

```



**Research quesion 3:**

```{r}

```

