---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
    toc: yes
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(ggmosaic)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```

* * *

## Part 1: Data

Genetics,  
* * *

## Part 2: Research questions

**Research quesion 1:**

Is the level of education of working people related to higher income?

**Research quesion 2:**

Does somehow people's weight related to vegetables consumption? 

**Research quesion 3:**

Is self employed mans who sleeps less than it is necessary, drink more alcohol than employed who sleeps enough?

* * *

## Part 3: Exploratory data analysis



### Research question 1: Is the level of education of working people related to higher income?

Let's dive into Codebook in order to select variables that can help us answer this question.

For education level the Codebook offers two variables: **educa** and calculated variable **\_educag**. While **educa** has six levels 

```{r education-levels}
levels(brfss2013$educa)
```

**\_educag** has four (in it's internals R replaces **\_educag** with **X_educag**, so we will refer the latter)

```{r x-education-levels}
levels(brfss2013$X_educag)
```

This four levels of educations is enough to understand the problem, so we will use **\_educag** in our research.  

Similary, let's check income variable. The Codebook offers two variables: **income2** and calculated variable**\_incomg**. 

```{r income-levels}
levels(brfss2013$income2)
```

```{r x-income-levels}
levels(brfss2013$X_incomg)
```

**X_incomeg** seems detailed enough so we will use it.

Let's take a closer look at variables. 

First of all, let's extract variables that we are interested into separate dataframe in order to work with less data volume to speedup calculations and do not mutate original dataframe if we will need.

```{r create-education-income-data}
education.income <- data.frame(income=brfss2013$X_incomg, education=brfss2013$X_educag)
head(education.income)
```

Let's check varaibles on the subject of NA (and related garbage) values.

```{r check-education-na}
education.income %>%
  group_by(education) %>%
  summarize(count=n())
```

```{r}
education.income %>%
  group_by(income) %>%
  summarize(count=n())
```

As we can see, there is NA values. They are useless in the context of current research, so let's get rid of them.

```{r}
education.income <- na.omit(education.income)
```

```{r}
education.income %>%
  group_by(education) %>%
  summarize(count=n())
```

```{r}
education.income %>%
  group_by(income) %>%
  summarize(count=n())
```

Now, when we have clean dataset, it's time to take a look at distribution of each variable.

```{r}
education.income %>%
  group_by(education) %>%
  summarize(freq=sprintf("%.2f %%", n()/nrow(education.income) * 100))
```

```{r}
ggplot(education.income, aes(x=education)) + geom_bar() + coord_flip()
```
 
 From the frequency table and bar plot above we can learn that the fraction of those who has no formal education is really small and is near 8%. The next two groups are nearly equal, the fraction of those who graduated high school and those who attended college or technical school are about 28% each. If we combine those groups we will see that the number of people who graduated high school, but didn't graduate college is dominant and equal 56%. 
 
```{r}
education.income %>%
  group_by(income) %>%
  summarize(freq=sprintf("%.2f %%", n()/nrow(education.income) * 100))
```
 
```{r}
ggplot(education.income, aes(x=income)) + geom_bar() + coord_flip()
```

```{r}
table.education.margin <- table(education.income$education, education.income$income) %>% prop.table(1)
colnames(table.education.margin) <- c('< $15K', '$15K-$25K', '$25K-$35K', '$35K-$50K', '$50K >')
table.education.margin
```

```{r}
table.income.margin <- table(education.income$education, education.income$income) %>% prop.table(2)
colnames(table.income.margin) <- c('< $15K', '$15K-$25K', '$25K-$35K', '$35K-$50K', '$50K >')
table.income.margin
```

```{r}
ggplot(education.income, aes(x=education, fill=income)) + geom_bar() + scale_x_discrete(labels = abbreviate) 
```

```{r}
ggplot(education.income, aes(x=education, fill=income)) + geom_bar(position='fill') + scale_x_discrete(labels = abbreviate) 
```

```{r}
ggplot(education.income) + geom_mosaic(aes(x = product(education), fill=income)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r echo=TRUE}
education.income$education <- recode(education.income$education, "Attended college or technical school" = "Graduated high school")
```

```{r}
education.income %>%
  group_by(education) %>%
  summarize(freq=sprintf("%.2f %%", n()/nrow(education.income) * 100))
```

```{r}
ggplot(education.income, aes(x=education)) + geom_bar() + coord_flip()
```

```{r echo=TRUE}
education.income$income <- recode(education.income$income, "$15,000 to less than $25,000" = "$15,000 to less than $50,000",
                                                            "$25,000 to less than $35,000" = "$15,000 to less than $50,000",
                                                            "$35,000 to less than $50,000" = "$15,000 to less than $50,000")
```

```{r}
education.income %>%
  group_by(income) %>%
  summarize(freq=sprintf("%.2f %%", n()/nrow(education.income) * 100))
```

```{r}
ggplot(education.income, aes(x=income)) + geom_bar() + coord_flip()
```

```{r}
ggplot(education.income, aes(x=education, fill=income)) + geom_bar(position='fill') + scale_x_discrete(labels = abbreviate) 
```

```{r}
table.education.margin <- table(education.income$education, education.income$income) %>% prop.table(1)
colnames(table.education.margin) <- c('< $15K', '$15K-$50K', '$50K >')
table.education.margin
```

```{r}
ggplot(education.income) + geom_mosaic(aes(x = product(education), fill=income)) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### Research question 2: Does somehow people's weight related to vegetables consumption?

In order to answer this question let's look at the variables that the codebook can provide. As for weight, there is a reference to **weight2** reported weight in pounds (without shoes). But this variable has mixed pounds/kilograms values (kilograms are used for values bigger than 9000), so we should preprocess it to achieve monotony. For our happiness dataset compilers already did this job and provide us with **wtkg3** - computed weight in kilograms. Let's explore it. Note, as codebook states, there is 2 implied decimal places, so the value 9754 kg we should perceive as 97,54 kg.

First of all, let's extract variables that we are interested into separate dataframe in order to work with less data volume to speedup calculations and do not mutate original dataframe if we will need. Also let's remove any NA from new data frame.

```{r create-weight-vegetables-data}
weight.vegetables <- data.frame(weight=brfss2013$wtkg3, vegetables=brfss2013$X_vegesum) %>% na.omit()
head(weight.vegetables)
```

```{r}
anyNA(weight.vegetables)
```

At first, let's explore 'weight' variable.

```{r}
weight <- weight.vegetables$weight
```

```{r}
str(weight)
```

It's a descrete numeric variable. 

```{r}
range(weight)
```

With range of values from 1 (person with weight of 0,01 kilogramm???) to 290,3 kilograms.

Let's say a couple of words about central tendency.

```{r}
mean(weight)
```

Mean roughly equal to 80 kg.

```{r}
median(weight)
```

And median roughly equal to 77 kg.

As the mean greater than median, we can suggest it's the right skewed distribution.

As for dispersion:

```{r}
paste(
  'Variance: ', round(var(weight)), '\n',
  'Standard deviation: ', round(sd(weight)), '\n',
  'IQR: ', round(IQR(weight)),
  sep = ''
) %>% cat
```

Very dispersed dataset with dense IQR.

Let's take a look at how the distribution is looks like.

```{r}
ggplot(data.frame(weight), aes(x=weight)) + geom_histogram()
```

```{r}
ggplot(data.frame(weight), aes(y=weight)) + stat_boxplot(geom = 'errorbar', width = 0.1) + geom_boxplot() + coord_flip()
```

By looking at graphs we can confirm the suggestion that the distribution skewed to the right. But also we can note that while the IQR is condensed, there is a lot of outliers. What to do with them? By looking at box plot we can suggest there is two types of outliers - weights less than ~40 kg and greater than ~125 kg. By the intuition weights of, for example, 200 kg, or 30 kg, is unusual weights and we can get rid of it if there is just a few records of such kinds of weights. First of all, let's find quartiles Q1 and Q3 to define borders of IQR.

```{r}
quantile(weight)
```

```{r}
q1 <- 6577
q3 <- 9072
iqr <- IQR(weight)
```

A commonly used rule says that low outliers are below Q1-1.5\*IQR and high outliers are above Q3+1.5\*IQR, so lets find those borders.

```{r}
outliers.border.low <- q1 - 1.5 * iqr
outliers.border.high <- q3 + 1.5 * iqr
```

And check how many outliers we have.

```{r}
length(weight[weight < outliers.border.low | weight > outliers.border.high])
```

```{r}
10471/length(weight)
```

Not so much, just about 2%. Let's go futher without them and remove outliers from variable 'weight' and from dataset 'weight.vegetables'.

```{r}
weight <- weight[weight >= outliers.border.low & weight <= outliers.border.high]
```

```{r}
weight.vegetables <- subset(weight.vegetables, weight >= outliers.border.low & weight <= outliers.border.high)
```

Let's check how distribution whithout outliers is looks like.

```{r}
ggplot(data.frame(weight), aes(x=weight)) + geom_histogram(binwidth = 1000)
```

Unimodal, slightly skewed to the right.

```{r}
ggplot(data.frame(weight), aes(y=weight)) + stat_boxplot(geom = 'errorbar', width = 0.1) + geom_boxplot() + coord_flip()
```

At second, let's explore 'vegetables' variable.

```{r}
vegetables <- weight.vegetables$vegetables
```

```{r}
str(vegetables)
```

It's a descrete numeric variable. 

```{r}
range(vegetables)
```

With range of values from 0 (no vegetables consumption) to 19827 (???).

Let's say a couple of words about central tendency.

```{r}
mean(vegetables)
```

```{r}
median(vegetables)
```

As the mean greater than median, we can suggest it's the right skewed distribution.

As for dispersion:

```{r}
paste(
  'Variance: ', round(var(vegetables)), '\n',
  'Standard deviation: ', round(sd(vegetables)), '\n',
  'IQR: ', round(IQR(vegetables)),
  sep = ''
) %>% cat
```

Very dispersed dataset with dense IQR.

Let's take a look at how the distribution is looks like.

```{r}
ggplot(data.frame(vegetables), aes(x=vegetables)) + geom_histogram()
```

```{r}
ggplot(data.frame(vegetables), aes(y=vegetables)) + stat_boxplot(geom = 'errorbar', width = 0.1) + geom_boxplot() + coord_flip()
```

By looking at graphs we can confirm the suggestion that the distribution skewed to the right. But also we can note that while the IQR is condensed, there is a lot of outliers. What to do with them? Let's explore if we can get rid of them. First of all, let's find quartiles Q1 and Q3 to define borders of IQR.

```{r}
quantile(vegetables)
```

```{r}
q1 <- 104
q3 <- 243
iqr <- IQR(vegetables)
```

A commonly used rule says that low outliers are below Q1-1.5\*IQR and high outliers are above Q3+1.5\*IQR, so lets find those borders.

```{r}
outliers.border.low <- q1 - 1.5 * iqr
outliers.border.high <- q3 + 1.5 * iqr
```

And check how many outliers we have.

```{r}
length(vegetables[vegetables < outliers.border.low | vegetables > outliers.border.high])
```

```{r}
10471/length(vegetables)
```

Not so much, just about 2%. Let's go futher without them and remove outliers from variable 'vegetablest' and from dataset 'weight.vegetables'.

```{r}
vegetables <- vegetables[vegetables >= outliers.border.low & vegetables <= outliers.border.high]
```

```{r}
weight.vegetables <- subset(weight.vegetables, vegetables >= outliers.border.low & vegetables <= outliers.border.high)
```

Let's check how distribution whithout outliers is looks like.

```{r}
ggplot(data.frame(vegetables), aes(x=vegetables)) + geom_histogram()
```

Unimodal, slightly skewed to the right.

```{r}
ggplot(data.frame(vegetables), aes(y=vegetables)) + stat_boxplot(geom = 'errorbar', width = 0.1) + geom_boxplot() + coord_flip()
```

Now, let's try to answer to the research question, is there any relation between weight and vegetables consumption? 

```{r}
ggplot(weight.vegetables, aes(x=vegetables, y=weight)) + geom_point()
```

Wow. The scatterplot is so dense, so it's absolutely impossible to find any kind of relation here. We can make things a bit clearer by reducing sample size and   

```{r}
ggplot(sample_n(weight.vegetables, 10000), aes(x=vegetables, y=weight)) + geom_point(shape=1, alpha = 0.5)
```

While there is recognized some 'cloud' with high density, still nothing. Let's add regression line to the plot.

```{r}
ggplot(weight.vegetables, aes(x=vegetables, y=weight)) + geom_point(shape=1, alpha = 0.5) + geom_smooth(method='lm', se = FALSE) 
```

Now we can see that there is small negative relation between weight and vegetables consumption. So there is a small chance the lighter people are, the more they eat vegetables. But for confidence, let's check the coorelation coefficient between weight and vegetables.  

```{r}
cor(weight.vegetables)
```

This table confirms the suggestion about small negative relation. 

**Research quesion 3: Is self employed mans who sleeps less than it is necessary, drink more alcohol than employed mans with normal sleep?**

The main idea behind this investigation is the next: self employed who sleeps less are working hard and so under pressure and they need to releive stress, while employed with normal sleep don't.    

Let's find in codebook variables we can use to make this research. We will look for something about sex, employment status, sleep hours and alcohol consumption. 

With the sex all is simple, it's 'sex' variable.

```{r}
str(brfss2013$sex)
```

```{r}
brfss2013 %>%
  group_by(sex) %>%
  summarize(count=n())
```

As for employment status, it's 'employ1' variable. Take a look at it quickly. 

```{r}
str(brfss2013$employ1)
```

```{r}
levels(brfss2013$employ1)
```

```{r}
brfss2013 %>%
  group_by(employ1) %>%
  summarize(count=n())
```

It's have enough data to split employed for wages and self employed people, so we can use it in research.

For filtering peoples by sleep adequacy, we should use 'sleptim1' variable, because it's the only one about sleep hours. In questionary, the question about sleep is looks the next: 'On average, how many hours of sleep do you get in a 24-hour period?', so we should decide ourselves how to split adequate/inadequate hours, but now let's just take a look at what the variable is.

```{r}
str(brfss2013$sleptim1)
```

```{r}
brfss2013 %>%
  group_by(sleptim1) %>%
  summarize(count=n())
```

Seems good enough to use in research. Go further.

As for alcohol consumption, the variable 'alcday5', Days In Past 30 Had Alcoholic Beverage, looks the most reasonable one, because other alcohol related variables have too much missing values. 

```{r}
str(brfss2013$alcday5)
```

```{r}
brfss2013 %>%
  group_by(alcday5) %>%
  summarize(count=n())
```

[101 - 199]	Days per week, [201 - 299]	Days in the past 30 days - so we will need to transform them to one view. 

Now, when all variables has been chosen, we will extract them into new data frame and will get rid of rows with any NA at the same time.

```{r}
employment.sleep.alcohol <- data.frame(
  sex=brfss2013$sex,
  employment=brfss2013$employ1,
  sleep=brfss2013$sleptim1,
  alcohol=brfss2013$alcday5
) %>% na.omit()

head(employment.sleep.alcohol)  
```

```{r}
anyNA(employment.sleep.alcohol)
```

Filter new dataset and drop sex column. 

```{r}
employment.sleep.alcohol <- employment.sleep.alcohol %>% filter(sex == 'Male' & employment %in% c('Self-employed', 'Employed for wages')) %>% select(-c(sex))
head(employment.sleep.alcohol) 
```

That's a moot point, what sleep we can consider as inadequate, so, based on the opionin the adult person should sleeps 7-9 hours per day, let's take 6 hours as an upper bound of the inadequate sleep.

```{r}
employment.sleep.alcohol <- employment.sleep.alcohol %>% mutate(sleep = if_else(sleep > 6, 'Adequate', 'Inadequate'))
head(employment.sleep.alcohol)
```

Calculate variable we are interested in and drop not needed.

```{r}
employment.sleep.alcohol <- employment.sleep.alcohol %>%
  filter((employment == 'Employed for wages' & sleep == 'Adequate') | (employment == 'Self-employed' & sleep == 'Inadequate'))
head(employment.sleep.alcohol)
```

Make 'alcohol' variable monotonous, transform it to drinking days per month, i.e, if value is 0, return it, if value is greater than or equal to 200, substract 200, if less than or equal to 107, subsctract 100, divide by 7 and multiply to 30.

```{r}
employment.sleep.alcohol <- employment.sleep.alcohol %>%
  mutate(alcohol = case_when(
    alcohol == 0 ~ 0,
    (alcohol >= 100 & alcohol <= 107) ~ round((alcohol - 100)/7 * 30),
    alcohol >= 200 ~ alcohol - 200
  )
)
head(employment.sleep.alcohol)
```

And now it's time to give an answer to the main question.

```{r}
ggplot(data.frame(employment.sleep.alcohol), aes(x=employment, y=alcohol)) + stat_boxplot(geom = 'errorbar', width = 0.1) + geom_boxplot() 
```

Wow, not the expected result. It seems there is no defference in alcohol consumption between self employed how sleeps little, and employed for wages who sleeps enough.