---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```

* * *

## Part 1: Data

Genetics,  
* * *

## Part 2: Research questions

**Research quesion 1:**

Is the level of education of working people lead to the higher income?

**Research quesion 2:**

Is the level of education of working native americans/alaskian people's lead to the higher income?

**Research quesion 3:**

Is childfree educated womens become to drink more alcohol with ages?

**Research quesion 4:**

Is self employed mans who sleeps less than it is necessary, drink more alcohol?

* * *

## Part 3: Exploratory data analysis

### Research question 1: Is the level of education of working people lead to the higher income?


**Research quesion 2:**

```{r}

```

**Research quesion 3:**

```{r}

```